Safe Transferable Regions

### 2016-07-03 paper/type-system.tex
Comments from DV

@@

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
\FB, hence it has no region annotations.
% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
defines a type rule for such values, allowing them to be typed.
The following type safety theorem shows that a well-typed program will never
attempt to dereference an ``invalid'' reference (a reference to an object
in a region that has been transferred or freed):
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}

>>>>

\dv{Why do we have the type parameters of $a$ and $b$ extending \C{Object} versus
  the actual class extending \C{Object<$\rho^a$>}? Also, instead of explicitly introducing
  all those $\rho^a$ variables, wouldn't it be nicer if we overloaded \C{this} to be a valid region,
  meaning {\bf the} region of this object? I think notationally that might be more lightweight?}

@@

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}

\dv{Note that we slightly changed notation here compared to the introduction, which
  had introduded n\C{new C@r} for new object allocations. I think we should make the
  connection in the beginning of this section.}

In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
\FB, hence it has no region annotations.
% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
defines a type rule for such values, allowing them to be typed.
Type safety theorem is now stated thus:
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}
